Spatial Reasoning and Statistical Graphics

### 2014-11-17 Paper/Draft1/TestingVisualAptitude.Rnw
Re-do explanation of lineup-type PCA, with additional explanation of future work

From figure~\ref{fig:lineupblockpca} we see that again, PC1 accounts for general visual aptitude. PC2, however, emphasizes the relationship between figure classification and Lineup task 1 (and to a lesser extent, on lineup task 2). PC3 differentiates the lineup task 3 from the visual search task; other tasks do not impact PC3 to a meaningful degree. PC4 separates lineup test 2 from the paper folding and card rotation tests and emphasizes the overlap with the visual search test. PC5 separates lineup test 2 and the figure classification test from the overlap between the other lineup tests and the visual search test. PC6 and PC7 account for \newtext{joint total of less than 10\% of the overall variability, and do not display any distinct patterns in the loadings.} %\Sexpr{as.numeric(round(100*importance[6], 1))}\% and \Sexpr{as.numeric(round(100*importance[7], 1))}\% of the variance, respectively; PC6 seems to differentiate the first two lineup tasks from the overlap between the third lineup task and the figure classification task.

\comment{What is the overall conclusion?}

\subsection{Lineup Plot Types}
\comment{Susan, would you agree in the assessment that the results from this section are inconclusive at best? Unless I am overlooking anything, I don't see substantial findings. We could move this section to the appendix, and just refer to it as being inconclusive.}
\newtext{
We can also compare participants' performance on specific types of lineup plots compared with their scores on the visual aptitude tests, for instance, accuracy on lineups which require mental rotation may be related to performance on the card rotation task. 
}

<<lineup-types,echo=F, fig.cap="Plot of scaled scores for different types of lineups\\label{fig:plottypesdensity}", fig.width=4, fig.height=8, out.width=".9\\linewidth", warning=F, message=F>>=
lineup.type.sum2$plot.type <- str_replace(lineup.type.sum2$plot.type, "rotated", "rot")
lineup.type.sum2$plot.type <- str_replace(as.character(lineup.type.sum2$plot.type), "error", "err")
lineup.type.sum2$plot.type <- factor(lineup.type.sum2$plot.type, 
                                     levels=c("violinplot", "jitter.boxplot", "bee.swarm.boxplot", "boxplot", 
                                              "scatterplot", "density", "histogram", 
                                              "qq.line", "qq.line.err", "qq.line.err.rot"), 
                                     labels=c("Violin Plot", "Boxplot + Points", "Bee Swarm Boxplot", "Boxplot", 
                                              "Scatterplot", "Density Plot", "Overlaid Histogram", 
                                              "QQ plot + Line", "QQ plot +\nAcceptance Band", 
                                              "Rotated QQ plot +\nAcceptance Band")
                                     )
qplot(data=subset(lineup.type.sum2, !is.na(plot.type)& testtype=="lineup"), 
      x=score, y=..scaled.., colour=factor(testnum), geom="density", cut=4) + 
  facet_wrap(~plot.type, ncol=2) + 
  xlab("Proportion of Lineups Correct") + 
  ylab("Density") + 
  geom_vline(aes(xintercept=0), linetype=3) + 
  scale_colour_discrete("Lineup Test Number") + 
  scale_x_continuous(breaks=c(-.33, 0, .33, .67, 1)) +
  scale_y_continuous(breaks=c(0, .5, 1)) +
  theme_bw() + 
  theme(legend.position="bottom", legend.direction="horizontal", 
        axis.text.x=element_text(angle=30, hjust=1),
        plot.margin=unit(c(0,0,0,0),"mm"), 
        axis.ticks.y=element_blank(), axis.text.y=element_blank())
@
\newtext{
Figure \ref{fig:plottypesdensity} compares performance on each different type of plot. As two different lineup tasks utilized boxplots to test different qualities of the distribution of data (outliers vs. difference in medians), different tasks are shown as different colors, so that accuracy on tasks which are shown in blue can be compared to other blue density curves. 
}

\newtext{
Figure \ref{fig:scatterplottypes} shows the association between scaled score on each type of lineup and score on the visual reasoning tests. Sample size for each plot type is fairly small - between 5 and 10 plots per individual, so there is low power for systematic inference, but we can establish that the card rotation task is much more significantly associated with the QQ plots tasks compared to the other tasks. In addition, rotated qq plots seem to be much more asssociated with the paper folding task scores than other qq plot tasks; this may be because they require more visual manipulation than other qq plots?
}
<<lineup-types-scores,echo=F, fig.cap="Plot of scaled lineup scores by aptitude test scores.\\label{fig:scatterplottypes}", fig.width=6, fig.height=12, out.width="\\linewidth", fig.env="figure", warning=F, message=F>>=
lineup.type.sum3 <- dcast(lineup.type.sum2, id + testtype ~ plot.type, value.var="score", fun.aggregate = mean)
other.scores <- subset(lineup.type.sum3, testtype!="lineup")[,c(1,2,13)]
names(other.scores)[3] <- "testscore"
lineup.scores <- melt(subset(lineup.type.sum3, testtype=="lineup")[,c(1,3:12)], id.vars = 1)
names(lineup.scores)[2:3] <- c("lineup.type", "lineup.pct")
lineup.type.sum3 <- merge(other.scores, lineup.scores, all.x=T, all.y=T)
lineup.type.sum3$lineup.type <- factor(lineup.type.sum3$lineup.type, 
                                     levels=c("Violin Plot", "Boxplot + Points", "Bee Swarm Boxplot", "Boxplot", 
                                              "Scatterplot", "Density Plot", "Overlaid Histogram", 
                                              "QQ plot + Line", "QQ plot +\nAcceptance Band", 
                                              "Rotated QQ plot +\nAcceptance Band"), 
                                     labels=c("Violin Plot", "Boxplot\n + Points", "Bee Swarm \nBoxplot", "Boxplot", 
                                              "Scatter\nPlot", "Density\nPlot", "Overlaid\nHistogram", 
                                              "QQ plot\n+ Line", "QQ plot\n+ Bands", 
                                              "Rotated QQ\nplot + Bands")
                                     )

cor.labels <- ddply(lineup.type.sum3, .(lineup.type, testtype), summarize, x=min(testscore), y=max(lineup.pct), label=paste0("cor = ", round(cor(testscore, lineup.pct), 2)))

>>>>

From figure~\ref{fig:lineupblockpca} we see that again, PC1 accounts for general visual aptitude. 
% PC2 emphasizes the relationship between figure classification and lineup task 1 (and to a lesser extent, on lineup task 2). PC3 differentiates the lineup task 3 from the visual search task; other tasks do not impact PC3 to a meaningful degree. PC4 separates lineup test 2 from the paper folding and card rotation tests and emphasizes the overlap with the visual search test. PC5 separates lineup test 2 and the figure classification test from the overlap between the other lineup tests and the visual search test. PC6 and PC7 account for a joint total of less than 10\% of the overall variability, and do not display any distinct patterns in the loadings.
\newtext{PC2 emphasizes the effect of performance on lineup test 1, which is associated with all of the measures of visual aptitude. PC3 emphasizes the association between performance on lineup test 3 and the visual search test, while PC4 emphasizes the association between performance on lineup test 2 and performance on the paper folding test. All three lineups, plus the figure classification, paper folding, and visual search tests contribute to PC5. PC6 and PC7 jointly account for less than 10\% of the variance in the data and do not display any distinct patterns in the loadings. }

\comment{If PC6-7 aren't important, then we still only have 5 PCs, or one additional PC from the grouped analysis. }
%\Sexpr{as.numeric(round(100*importance[6], 1))}\% and \Sexpr{as.numeric(round(100*importance[7], 1))}\% of the variance, respectively; PC6 seems to differentiate the first two lineup tasks from the overlap between the third lineup task and the figure classification task.

\subsection{Lineup Plot Types}
\comment{Susan, would you agree in the assessment that the results from this section are inconclusive at best? Unless I am overlooking anything, I don't see substantial findings. We could move this section to the appendix, and just refer to it as being inconclusive.}
% \comment{What is the overall conclusion?}
\newtext{
We can also compare participants' performance on specific types of lineup plots compared with their scores on the visual aptitude tests, for instance, accuracy on lineups which require mental rotation may be related to performance on the card rotation task. 
}

<<lineup-types,echo=F, fig.cap="Plot of scaled scores for different types of lineups\\label{fig:plottypesdensity}", fig.width=4, fig.height=8, out.width=".9\\linewidth", warning=F, message=F>>=
lineup.type.sum2$plot.type <- str_replace(lineup.type.sum2$plot.type, "rotated", "rot")
lineup.type.sum2$plot.type <- str_replace(as.character(lineup.type.sum2$plot.type), "error", "err")
lineup.type.sum2$plot.type <- factor(lineup.type.sum2$plot.type, 
                                     levels=c("violinplot", "jitter.boxplot", "bee.swarm.boxplot", "boxplot", 
                                              "scatterplot", "density", "histogram", 
                                              "qq.line", "qq.line.err", "qq.line.err.rot"), 
                                     labels=c("Violin Plot", "Boxplot + Points", "Bee Swarm Boxplot", "Boxplot", 
                                              "Scatterplot", "Density Plot", "Overlaid Histogram", 
                                              "QQ plot + Line", "QQ plot +\nAcceptance Band", 
                                              "Rotated QQ plot +\nAcceptance Band")
                                     )
qplot(data=subset(lineup.type.sum2, !is.na(plot.type)& testtype=="lineup"), 
      x=score, y=..scaled.., colour=factor(testnum), geom="density", cut=4) + 
  facet_wrap(~plot.type, ncol=2) + 
  xlab("Proportion of Lineups Correct") + 
  ylab("Density") + 
  geom_vline(aes(xintercept=0), linetype=3) + 
  scale_colour_discrete("Lineup Test Number") + 
  scale_x_continuous(breaks=c(-.33, 0, .33, .67, 1)) +
  scale_y_continuous(breaks=c(0, .5, 1)) +
  theme_bw() + 
  theme(legend.position="bottom", legend.direction="horizontal", 
        axis.text.x=element_text(angle=30, hjust=1),
        plot.margin=unit(c(0,0,0,0),"mm"), 
        axis.ticks.y=element_blank(), axis.text.y=element_blank())
@
\newtext{
Figure \ref{fig:plottypesdensity} compares performance on each different type of plot. As two different lineup tasks utilized boxplots to test different qualities of the distribution of data (outliers vs. difference in medians), different tasks are shown as different colors, so that accuracy on tasks which are shown in blue can be compared to other blue density curves. 
}
While lineups constitute a distinct principal component when aggregated into a single score, this PCA of the separate lineup types and the cognitive tests indicates that different lineup experiments exist in different principal component loadings. Overall, there is an additional principal component gained from separating the lineup blocks by experiment type. As lineup tasks 1 and 2 contained similar plot types, it is possible that those two tasks overlap in the component space while lineup task 3 is distinct. }

\newtext{
Figure \ref{fig:scatterplottypes} shows the association between scaled score on each type of lineup and score on the visual reasoning tests. Sample size for each plot type is fairly small - between 5 and 10 plots per individual, so there is low power for systematic inference, but we can establish that the card rotation task is much more significantly associated with the QQ plots tasks compared to the other tasks. In addition, rotated qq plots seem to be much more asssociated with the paper folding task scores than other qq plot tasks; this may be because they require more visual manipulation than other qq plots?
}
<<lineup-types-scores,echo=F, fig.cap="Plot of scaled lineup scores by aptitude test scores.\\label{fig:scatterplottypes}", fig.width=6, fig.height=12, out.width="\\linewidth", fig.env="figure", warning=F, message=F>>=
lineup.type.sum3 <- dcast(lineup.type.sum2, id + testtype ~ plot.type, value.var="score", fun.aggregate = mean)
other.scores <- subset(lineup.type.sum3, testtype!="lineup")[,c(1,2,13)]
names(other.scores)[3] <- "testscore"
lineup.scores <- melt(subset(lineup.type.sum3, testtype=="lineup")[,c(1,3:12)], id.vars = 1)
names(lineup.scores)[2:3] <- c("lineup.type", "lineup.pct")
lineup.type.sum3 <- merge(other.scores, lineup.scores, all.x=T, all.y=T)
lineup.type.sum3$lineup.type <- factor(lineup.type.sum3$lineup.type, 
                                     levels=c("Violin Plot", "Boxplot + Points", "Bee Swarm Boxplot", "Boxplot", 
                                              "Scatterplot", "Density Plot", "Overlaid Histogram", 
                                              "QQ plot + Line", "QQ plot +\nAcceptance Band", 
                                              "Rotated QQ plot +\nAcceptance Band"), 
                                     labels=c("Violin Plot", "Boxplot\n + Points", "Bee Swarm \nBoxplot", "Boxplot", 
                                              "Scatter\nPlot", "Density\nPlot", "Overlaid\nHistogram", 
                                              "QQ plot\n+ Line", "QQ plot\n+ Bands", 
                                              "Rotated QQ\nplot + Bands")
                                     )
\newtext{The relationship between participant performance on different types of lineups (and different types of plots) and performance on tests of spatial ability bears further investigation; this study suggests that there may be an effect, but there is simply not enough variation to make definitive conclusions about the relationship between different measures of visual aptitude and performance on specific lineup tasks. }

cor.labels <- ddply(lineup.type.sum3, .(lineup.type, testtype), summarize, x=min(testscore), y=max(lineup.pct), label=paste0("cor = ", round(cor(testscore, lineup.pct), 2)))
\newtext{A larger study could not only address the question of which lineup tasks require certain visual skills, but could also address the use of different types of plots from a perceptual perspective. A preliminary analysis of performance on different types of plots is available in appendix \ref{app:plottypes}, but a larger study is needed for definitive results. The advantage of the lineup protocol is that it allows us to not only consider individual performance but also to compare aggregate performance on different types of plots. Integrating information about the visual skills required for each type of plot provides information about the underlying perceptual skills and experience required to read different types of plots.}
